A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-18 tex/body.tex
reorder

\paragraph{Phase 2: Applying Acyclicity and Uniqueness.}
%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
The premises of the rules that make up the second phase refer to a mapping $\Val$
that associates a $\mu$-term as representative with each equivalence class.
%Recall that $\mu$-bindings can be used for representing possibly cyclic terms and values.
%(Section~\ref{sec:introduction}).
% for example, $\MU x.\; \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
A $\mu$-term $t$ of type $\tau$ describes a class of values of
$\tau$ that terms in that equivalence class can take in the models of $\Ec$.
%either in the case when $\tau$ is a codatatype or datatype type,
When $\tau$ is a datatype, % \in \Data$,
a cyclic $t$ describes an infeasible class of values.
%$\tau \not\in \Codata$.

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or a node of the form
$\MU x.\: \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable $x$ need not occur freely in the binder's body, in which case
the binder can be omitted.
In the \rn{Acyclic} rule,
$\FV( u )$ denotes the variables occurring freely in~$u$.
%%% But there are no constants of type $\tau \in \Nondata$ -- at most variables!
%
%For uniformity, we also consider $%\MU x.\;
%\const{c}$ to be a $\mu$-term if $\const{c}$ is a constant of type $\tau \in \Nondata$.
%
%%% I think this should be clear enough from the "Conventions" paragraph at the
%%% end of the introduction.
%%
% which we define recursively:
%if $\const{C} : \vec \tau \rightarrow \upsilon$
%and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
%then $\MU x.\; \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
%
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. \hfill It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. \hfill The
\goodbreak %%% TYPESETTING: horrible hacks (including two hfills)
\noindent
notation $t \aequiv u$, for $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables\begin{rep}---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{C}( x ) \not\aequiv \MU y.\; \const{C}( \const{C}( y ) )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

The mapping $\Val$ is computed as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$
%not occurring in $\tEc$,
and set $\Val\ec{u} := {\Varec{u}}$. Therefore, there are initially no
constraints on the values for any equivalence class $\ec{u}$. The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the variable-capturing substitution of $t$ for
$x$ in $\Val$'s range. It is easy to see that the height of terms produced as a
result of the unfolding is bounded by the number of equivalence classes of
$\Ec$, and thus the construction of $\Val$ will terminate.

@@

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C}, \const{D}, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
w \teq \const{C}( z ),\;
z \teq \const{D}( x ),\;
u \teq \const{C}( x ),\;
v \teq \const{E}( y ),\;
x \not\teq z
\}$.
After applying the calculus to saturation on $\Ec$, the mapping $\Val$ is as follows:
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l}
\[\begin{array}{r@{}c@{}r@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}l@{\;}l}
\Val \ec{x} & = & {\Varec{x}} &
\Val \ec{z} & = & \MU {\Varec{z}}. & \multicolumn{2}{@{}l@{}}{\const{D}( {\Varec{x}} )} & &
\Val \ec{u} & {}\,=\,{} & \MU {\Varec{u}}. & \const{C}( {\Varec{x}} ) \\
\Val \ec{y} & = & {\Varec{y}} &
\Val \ec{w} & {}\,=\,{} & \MU {\Varec{w}}. & \const{C}( & \MU {\Varec{z}}. & \const{D}( {\Varec{x}} ) ) &
\Val \ec{v} & = & \MU {\Varec{v}}. & \const{E}( {\Varec{y}} )
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, namely ${\Varec{x}}$ and ${\Varec{y}}$.
Modulo $\alpha$-equivalence, the set \smash{$\mathcal{T}^{\smash{\J}}_{{\Varec{x}}}( \Val )$} is
$%\begin{array}{c}
\{
\MU x.\; \const{D}( x ),\;
\MU x.\; \const{C}( \const{D}( x )),\allowbreak\;
\const{C}( \MU x.\; \const{D}( x )),\;
\MU x.\; \const{C}( x )
\}
%\end{array}
$.
Consider a normal interpretation $\J$ that interprets variables in $\Ec$ based on the mapping $\Val$:
$x$ is interpreted as $\Val \ec{x}$, $y$ as $\Val \ec{y}$, and so on.
Assigning a value for $\Val \ec{x}$ that is $\alpha$-equivalent to a value in $\mathcalx{T}^{\smash{\J}}_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\Ec$ to be falsified by $\J$.
For example, assign $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{z}$ is $\MU {\Varec{z}}.\; \const{D}( \MU {\Varec{x}}.\; \const{D}( {\Varec{x}} ) )$,
which has normal form $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq z$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{x}}$ is fresh, %not $\alpha$-equivalent to any term in $\mathcalx{T}^{\smash{\M}}_{{\Varec{x}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign a value such as
$\MU {\Varec{x}}.\; \const{E}( {\Varec{x}} )$ or
$\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for ${\Varec{x}}$ may cause the range of $\Val$ to contain abnormal terms.
For example, after assigning $\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$ to $\Varec{x}$, we have that
$\Val \ec{u}$ is $\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) ) )$,
which has normal form
$\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( {\Varec{u}} ) )$.
\xend
\end{examplex}

>>>>

\paragraph{Phase 2: Applying Acyclicity and Uniqueness.}
%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
The premises of the rules that make up the second phase refer to a mapping $\Val$
that associates a $\mu$-term with each equivalence class as its representative.
%Recall that $\mu$-bindings can be used for representing possibly cyclic terms and values.
%(Section~\ref{sec:introduction}).
% for example, $\MU x.\; \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
A $\mu$-term $t\typ{\tau}$ describes a class of $\tau$ values
that terms in that equivalence class can take in the models of $\Ec$.
%either in the case when $\tau$ is a codatatype or datatype type,
When $\tau$ is a datatype, % \in \Data$,
a cyclic $\mu$-term describes an infeasible class of values.
%$\tau \not\in \Codata$.

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or a node of the form
$\MU x.\: \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable $x$ need not occur freely in the $\mu$-binder's body, in which case
the binder can be omitted.
In the \rn{Acyclic} rule,
$\FV( u )$ denotes the variables occurring freely in~$u$.
%%% But there are no constants of type $\tau \in \Nondata$ -- at most variables!
%
%For uniformity, we also consider $%\MU x.\;
%\const{c}$ to be a $\mu$-term if $\const{c}$ is a constant of type $\tau \in \Nondata$.
%
%%% I think this should be clear enough from the "Conventions" paragraph at the
%%% end of the introduction.
%%
% which we define recursively:
%if $\const{C} : \vec \tau \rightarrow \upsilon$
%and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
%then $\MU x.\; \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
%
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. \,It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. \,The
\goodbreak %%% TYPESETTING: horrible hacks (including two \,'s)
\noindent
$\alpha$-\emph{equivalence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables\begin{rep}---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{C}( x ) \not\aequiv \MU y.\; \const{C}( \const{C}( y ) )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

The mapping $\Val$, on which the rules depends, is computed as follows.
%
With each equivalence class $\ec{u\typ{\tau}}$, we associate a fresh variable $\Varec{u}\typ{\tau}$
%of the same type as $u$
%not occurring in $\tEc$,
and set $\Val\ec{u} := {\Varec{u}}$. Therefore, there are initially no
constraints on the values for any equivalence class $\ec{u}$. The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the \emph{variable-capturing} substitution of $t$ for
$x$ in $\Val$'s range. It is easy to see that the height of terms produced as a
result of the unfolding is bounded by the number of equivalence classes of
$\Ec$, and thus the construction of $\Val$ will terminate.

@@

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C},\, \const{D},\, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
w \teq \const{C}( z ),\;
z \teq \const{D}( x ),\;
u \teq \const{C}( x ),\;
v \teq \const{E}( y ),\;
x \not\teq z
\}$.
After applying the calculus to saturation on $\Ec$, the mapping $\Val$ is as follows:
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l}
\[\begin{array}{r@{}c@{}r@{}l@{\qquad}r@{}c@{}r@{}l@{\qquad}r@{}c@{}l}
\Val \ec{u} & {}\,=\,{} & \MU {\Varec{u}}.\; & \const{C}( {\Varec{x}} ) &
\Val \ec{v} & = & \MU {\Varec{v}}.\; & \const{E}( {\Varec{y}} ) &
\Val \ec{y} & {}\,=\,{} & {\Varec{y}} \\
\Val \ec{z} & = & \MU {\Varec{z}}.\; & \const{D}( {\Varec{x}} ) &
\Val \ec{w} & {}\,=\,{} & \MU {\Varec{w}}.\; & \const{C}( \MU {\Varec{z}}.\; \const{D}( {\Varec{x}} ) ) &
\Val \ec{x} & = & {\Varec{x}}
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, namely ${\Varec{x}}$ and ${\Varec{y}}$.
Modulo $\alpha$-equivalence, the set \smash{$\mathcal{T}^{\smash{\J}}_{{\Varec{x}}}( \Val )$} is
$%\begin{array}{c}
\{
\MU x.\; \const{D}( x ),\;
\MU x.\; \const{C}( \const{D}( x )),\allowbreak\;
\const{C}( \MU x.\; \const{D}( x )),\;
\MU x.\; \const{C}( x )
\}
%\end{array}
$.
Consider a normal interpretation $\J$ that interprets variables in $\Ec$ based on the mapping $\Val$:
$x$ is interpreted as $\Val \ec{x}$, $y$ as $\Val \ec{y}$, and so on.
Assigning a value for $\Val \ec{x}$ that is $\alpha$-equivalent to a value in $\mathcalx{T}^{\smash{\J}}_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\Ec$ to be falsified by $\J$.
For example, assign $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{z}$ is $\MU {\Varec{z}}.\; \const{D}( \MU {\Varec{x}}.\; \const{D}( {\Varec{x}} ) )$,
which has normal form $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq z$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{x}}$ is fresh, %not $\alpha$-equivalent to any term in $\mathcalx{T}^{\smash{\M}}_{{\Varec{x}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign a value such as
$\MU {\Varec{x}}.\; \const{E}( {\Varec{x}} )$ or
$\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for ${\Varec{x}}$ may cause the range of $\Val$ to contain abnormal terms.
For example, after assigning $\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$ to $\Varec{x}$, we have that
$\Val \ec{u}$ is $\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) ) )$,
which has normal form
$\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( {\Varec{u}} ) )$.
\xend
\end{examplex}
